# Gaussian LM {#gen-model}

A Linear Model predicts a dependent (or response) variable, which is continuous and approximately normally distributed, from one or more independent (or predictor) variables. A normal statistical distribution, also referred to as a Gaussian distribution (after the brilliant German mathematician Carl Friedrich Gauss), assumes the data are drawn from a distribution that is symmetric and can be summarised by the arithmetic mean and standard deviation. Independent variables may also be continuous, categorical, or a combination of continuous and categorical. Categorical variables are commonly referred to as _factors_, which have a series of _levels._ For example, a factor might be sex, which has two levels, male and female.

A GLM comprises three components: 1. the linear predictor, which is a linear function of the predictor variable; 2. the conditional probability distribution of the response variable, which is the distribution of the response variable across the regression line for the given set of predictor variables; 3. the link function, which connects the linear predictor with the mean of the conditional probability distribution. 

Choice of conditional probability distribution (such as Gaussian, binomial, Bernoulli, Poisson, gamma, beta, etc.) is not based on the distribution of the raw response variable, but rather on variable characteristics, such as whether the variable is continuous or discrete, bounded or unbounded. Choice of conditional distribution largely determines which link function is most appropriate (such as identity, log, logit, inverse, etc.), though choice of link function can be refined as part of the model fitting process.

## European bitterling territoriality {#bitterling}

In this Chapter we fit a Linear Model with a Gaussian conditional distribution and an identity link function to a set of data on male European bitterling (_Rhodeus amarus_) territorial behaviour. Bitterling are small freshwater fish with an unusual breeding system. During the breeding season, male bitterling are aggressively territorial and guard freshwater mussels. Female bitterling develop a long egg-laying tube (‘ovipositor’) that they use to place their eggs in the gills of the mussel, which the males then fertilise.

A study was conducted in Lake Dědová near Lanžhot in the Czech Republic to measure the response distance of male bitterling to a rival when they were guarding a mussel. Male response distance was measured by gradually moving a model of a male bitterling towards a territorial male that was guarding a mussel. The response distance was the horizontal distance that it was possible to move the model towards a guarded mussel before the territorial male attacked it. After obtaining an estimate of the response distance, territorial males were captured with a hand net and their length measured, after which they were immediately released.

In addition, males were randomly allocated to a food supplement treatment, with approximately half the males in the study receiving a 1 g cube of freeze-dried _Tubifex_ worms daily for six days before the start of data recording. The remaining males received no food supplement, but did experience disturbance each day that was comparable to those receiving a food supplement.

First, we load required libraries (see Section \@ref(functions)) and import the data (see Section \@ref(import)).

*__Load libraries__*

```{r ch4_libraries, echo=TRUE, warning=FALSE, message=FALSE}
library(arm)
library(car)
library(dplyr)
library(ggplot2)
library(GGally)
library(lattice)
library(lawstat)
library(outliers)
library(tidyverse)
library(rlang)
library(gridExtra)
library(glmmTMB)
library(tidyverse)
library(DHARMa)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
```

*__Import data__*

Data for European bitterling are saved in a comma-separated values (CSV) file `bitter.csv` and are imported into a dataframe in R using:  

```{r ch4-import-bitt, echo=TRUE, warning=FALSE, message=FALSE}
bitt <- read.csv(file = "bitter.csv",                   
               header = TRUE,                            
                  dec = ".",                          
     stringsAsFactors = TRUE)
```

Start by inspecting dataframe `bitt`:

```{r ch4-str-bitt, comment = "", echo=TRUE, warning=FALSE, message=FALSE}
str(bitt, width = 50, strict.width = "cut")
```

The dataframe comprises `r nrow(bitt)` observations of `r ncol(bitt)` variables. Each row in the dataframe represents an observation for a different male bitterling (`male`). The variable `sl` is continuous and represents the standard length (in mm) of each male bitterling, while the variable `supp_feed` is categorical (though coded numerically) indicating those males that received no food supplement (`0`) and those that did (`1`).  The variable `resp_dist` is the aggressive response distance (in cm) and is the response (dependent) variable of interest.

The supplementary feeding covariate (`supp_feed`) is categorical but currently coded numerically. We need to designate this variable as a factor (`fSupp`):

```{r ch4-fact, comment = "", echo=TRUE, warning=FALSE, message=FALSE}
bitt$fSupp <- as.factor(bitt$supp_feed) 
```

For clarity, we recode `fSupp` as 'No' and 'Yes' instead of '0' and '1'.

```{r ch4-recode, comment = "", echo=TRUE, warning=FALSE, message=FALSE}
bitt$fSupp <- dplyr::recode(bitt$fSupp,
                           "0"="No",
                           "1"="Yes")  
levels(bitt$fSupp)
```

Let's also check for missing data.

```{r ch4-NAs, comment = "", echo=TRUE, warning=FALSE, message=FALSE}
colSums(is.na(bitt))
```

No missing data.

## Six steps to fitting a (G)LM {#GLM7steps4}

We will adopt a six-step protocol to fitting a (G)LM. These are:

**_1. State the question_**  
**_2. Perform a data exploration_**  
**_3. Select and fit a statistical model_**  
**_4. Conduct model checks_**  
**_5. Interpret and present model output_**  
**_6. Visualise the results_**  

### State the question

This study was conducted to understand the extent to which the territorial behaviour of male bitterling is a function of male size and body condition. Our predictions were that larger males would be more effective in responding to intruders to their territory than smaller males. A further prediction was that, given that territoriality is energetically expensive, and males are often constrained in their feeding while engaged in territory defence, supplementing the diets of males would also increase the aggressive response distance of males. A final prediction was that these two variables would interact; specifically that the effect of body size on response distance would be less pronounced in males that received a food supplement; i.e. energy depletion is more severe for larger males.

Consequently there are three specific predictions to test:

1. A positive association between male body size, measured as standard length (`sl`), and response distance (`resp_dist`).

2. A positive association between provision of supplementary food (`fSupp`) and response distance.

3. An interaction between male body size (`sl`) and supplementary feeding (`fSupp`) and response distance (`resp_dist`), with a steeper slope predicted between body size and response distance for males that did not receive supplementary food.  

### Perform a data exploration  

We start by conducting a data exploration to identify any potential problems with the data. 

*__Seven-step data exploration protocol__*  

We adopt a protocol comprising 7 steps that is intended to identify:  

**_1.	Outliers in response and independent variables_**  
**_2.	Normality and homogeneity of the response variable_**  
**_3.	Balance of categorical variables_**  
**_4.	An excess of zeros in the response variable_**  
**_5.	Multicollinearity among independent variables_**  
**_6.	Relationships among response and independent variables_**  
**_7.	Independence of the response variable_**  

#### Outliers in response and independent variables {#outliers4}  

An outlier is an observation that has a relatively large or small value compared to the other observations. (G)LMs are sensitive to the presence of outliers in data. Outliers are best identified visually. For the categorical variable `fSupp`, we use a boxplot to visualise how `resp_dist` varies with the provision of a food supplement.  

```{r ch4-My-theme, warning=FALSE, echo=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=50)}
My_theme <- theme(panel.background = element_blank(),  
          panel.border = element_rect(fill = NA, size = 1),  
          strip.background = element_rect(fill = "white",  
          color="white"), text=element_text(size = 12),  
          panel.grid.major = element_line(colour = "white"),  
          panel.grid.minor = element_line(colour = "white"))  
```

Then plot:  

(ref:ch4-box-supp) **Boxplot of response distances (cm) of territorial male bitterling that did and did not receive a food supplement.**  

```{r ch4-box-supp, fig.cap='(ref:ch4-box-supp)', fig.align='center', fig.dim=c(6, 4), warning=FALSE, echo=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=50)}
bitt %>%
  ggplot(aes(y = resp_dist, x = fSupp,  
                        group = fSupp)) +  
  labs(y="Response distance (cm)",  
       x="Supplement") +  
  geom_boxplot(fill = "gray88") +  
  My_theme
```

Fig. \@ref(fig:ch4-box-supp) visualises the median and spread of male response distance, with the median represented as a thick horizontal line and the 25% and 75% quartiles (the inter-quartile range or IQR) represented by the box. The vertical black lines extending from the box represent the range of the data. The boxplot shows that there was a greater average response distance for males that received a food supplement. The size of the IQR and range of the data are, however, similar, irrespective of whether males received a food supplement, indicating homogeneity in the data.  

To identify outliers for continuous variables we use multi-panel dotplots using the `multi_dotplot` function.

```{r ch4-dot-function, message = FALSE, echo=FALSE, warning=FALSE}
multi_dotplot <- function(filename, Xvar, Yvar){
  filename %>%
    ggplot(aes(x = {{Xvar}})) +
    geom_point(aes(y = {{Yvar}})) +
    My_theme +
    coord_flip() +
    labs(x = "Order of data")}
```


```{r ch4-order, message = FALSE, echo = FALSE, warning=FALSE}
bitt <- bitt %>%
  mutate(order = seq(1:nrow(bitt)))
```

Identify those continuous variables we wish to plot (`sl`, `resp_dist`):  

```{r ch4-grid1, message = FALSE, echo=TRUE, warning=FALSE}
p1 <- multi_dotplot(bitt, order, sl)        + My_theme
p2 <- multi_dotplot(bitt, order, resp_dist) + My_theme
```

Use the `grid.arrange` function from the `gridExtra` package to arrange these plots in our preferred format:  

(ref:ch4-dotplot1) **Dotplots of the continuous variables sl and resp_dist. Data are arranged by the order they appear in the dataframe (bottom to top).**  

```{r ch4-dotplot1, fig.cap='(ref:ch4-dotplot1)', message = FALSE, echo=TRUE, warning=FALSE}
grid.arrange(p1, p2, nrow = 1)
```

In Fig. \@ref(fig:ch4-dotplot1) the dotplots for `day` and `resp_dist` show no prominent outliers.  

#### Normality and homogeneity of the response variable {#normality4}  

An assumption of a Gaussian LM is that the response variable is normally distributed **at each covariate value**. The distribution of a continuous variable can be visualized by dividing the x-axis into “bins” and counting the number of observations in each bin as a frequency polygon using the `geom_freqpoly()` function from the `ggplot2` plotting package:  

```{r ch4-freqpoly, fig.cap='(ref:ch4-freqpoly)', message = FALSE, echo=TRUE, warning=FALSE}
bitt %>% ggplot(aes(resp_dist)) +
  geom_freqpoly(bins = 8) +
  labs(x = "Response distance (cm)", y = "Frequency") +
  My_theme
```

(ref:ch4-freqpoly) **Frequency polygon of the response distance of territorial male bitterling.**  

The frequency polygon plot of the dependent variable (Fig. \@ref(fig:ch4-freqpoly)) is broadly normal. However, this figure ignores the factors, such as male size (`sl`) or food supplement (`fSupp`), that may explain deviation from normality. At this stage, we will proceed with the data exploration bearing in mind that the raw data values for the dependent variable are broadly normally distributed.  

It is possible to test for normality using the Shapiro-Wilk test. The null hypothesis for this test is that the distribution of the data follows a normal distribution. If the P-value is below 0.05 (P <0.05) the null hypothesis is rejected and the conclusion is that the data depart from normality.  

```{r ch4-shapiro, message = FALSE, echo=FALSE, warning=FALSE}
shapiro.test(bitt$resp_dist)
```

The test shows no significant departure from normality (P = 0.766).  

_Homogeneity of variance_ is an even distribution of covariate values around the mean and is another important assumption of many statistical tests. Without homogeneity of variance estimated P-values are unreliable. There are several ways to measure homogeneity of variance.  

To visualise the homogeneity of the response variable in relation to a categorical covariate a boxplot is illustrative. Fig. \@ref(fig:ch4-box-supp) showed no obvious variation in the spread of response distance data between levels of the factor `fSupp`,  indicating homogeneity. A scatterplot is better to visualise homogeneity of variance in relation to a continuous covariate.  

(ref:ch4-scatter2-bitt) **Scatterplot of `resp_dist` and `sl`. The variance in response distance is similar across the range of male SLs, indicating homogeneity of the data.**  

```{r ch4-scatter2-bitt, fig.cap='(ref:ch4-scatter2-bitt)', fig.align='center', fig.dim=c(6, 4), message = FALSE, echo=FALSE, warning=FALSE}

bitt %>% 
  ggplot(aes(x = sl, y = resp_dist)) +
  geom_jitter(shape = 19, size = 3.5, height = 0.05, 
              width = 0.05, alpha = 0.5) +
  labs(x = "Male SL (mm)", y = "Response distance (cm)") +
  xlim(30,70) + ylim(75,250) +
  My_theme
```

The scatterplot (Fig. \@ref(fig:ch4-scatter2-bitt)) shows limited variability in male response distance across the full range of male SLs, indicating that the data probably do not deviate significantly from homogeneity.   

There are several tests of homogeneity of variance, should we need one, such as Bartlett's Test, the F-ratio test, and Levene's test. The first two of these tests assume normality of the data. If your data deviate from normality they should not be used. Levene's test does not assume normality. An alternative is the Brown & Forsythe test, which uses the median rather than mean in its estimation, and is robust to departures from normality. This test is based on Levene's test and can be obtained using the `levene.test()` function from the `lawstat` package:  

```{r ch4-levene-bitt, message=FALSE, echo=TRUE, warning=FALSE}
leveneTest(bitt$resp_dist,
           bitt$fSupp,
           location = c("median"), 
           trim.alpha = 0.25)
```

Which confirms the data do not deviate from homogeneity (P = 0.518).  

#### Balance of categorical variables {#balance4}    

Imbalance in data occurs when we have a categorical variable with markedly different numbers of observations of different levels. If a model contains one or more unbalanced variables, and particularly if variances among levels also differ, then fitted models have reduced power.  

Imbalance can be identified using the `table` command from base R.  

```{r ch4-table-supp, message=FALSE, echo=TRUE, warning=FALSE}
table(bitt$fSupp)
```

Balance between levels of the food supplement treatment is good.  

#### An excess of zeros in the response variable {#zeros4}   

Zeros should not be omitted from a dataset. However, an excess of zeros in the response variable can cause problems in an analysis, though there are a number of ways of dealing with 'zero inflation' in data. The first step is to identify whether there is a potential problem. The percentage of zeros in the response variable can be estimated as:  

```{r ch4-zeros, message=FALSE, echo=TRUE, warning=FALSE}
sum(bitt$resp_dist == 0, na.rm = TRUE) * 100 / nrow(bitt)
```

There are no zeros in the response variable for this dataset. Logically, a response distance by males of 0 cm would represent no response, so the fact that we have no zeros in this case is not unexpected.  

#### Multicollinearity among covariates  

Along with normality of residuals and homogeneity of variance, an additional assumption of linear modelling is independence of the independent variables. In ecological studies it is not unusual to collect a large number of variables, which are often highly correlated. If covariates in a model are correlated, then the model may produce unstable parameter estimates with inflated standard errors.  

Multicollinearity can be tested in several ways. We can obtain a comprehensive summary of the relationship between the two model covariates using the `ggpairs` command from the `GGally` package:  

(ref:ch4-ggpairs) **Plot matrix of bitterling standard length (mm) and supplementary feeding treatment. The top left panel shows a frequency plot of standard length split by feeding treatment, while the top right shows the same data expressed as a boxplot. The lower left panel shows a length-frequency histogram of standard lengths, with those for males that did not receive supplementary feeding above and those that did, below. The lower right panel shows the total number of individual males in each supplementary feeding treatment.**  

```{r ch4-ggpairs, fig.cap='(ref:ch4-ggpairs)', fig.align='center', fig.dim=c(6, 4), cache = TRUE, message = FALSE, echo=TRUE, warning=FALSE}
bitt %>% 
    ggpairs(columns = c("sl", "fSupp"), 
         aes(colour = fSupp, alpha = 0.8), 
              combo = wrap("facethist"))
```

The plot matrix in Fig. \@ref(fig:ch4-ggpairs) demonstrates no clear pattern of collinearity between the two covariates and illustrates good overlap in male standard length between levels of the (randomly assigned) feeding treatment.  

Another approach to identifying multicollinearity is by calculating a variance inflation factor (VIF) for each variable. The VIF is an estimate of the proportion of variance in one predictor explained by all the other predictors in the model. A VIF of 1 indicates no collinearity. VIF values above 1 indicate increasing degrees of collinearity. VIF values exceeding 3 are considered problematic Zuur _et al_. (2010), in which case the variable with the highest VIF should be removed from the model and the VIFs for the model recalculated.  

The VIF for a model can be estimated using the `vif` function from the `car` package:  

```{r ch4-vif, message=FALSE, echo=TRUE, warning=FALSE}
round(vif(lm(resp_dist ~ sl + fSupp, data = bitt)),2)
```

For the bitterling model the estimated VIFs are <3, so there is no problem with multicollinearity.  

#### Relationships among dependent and independent variables  

Visual inspection of the data using plots is a critical step and will illustrate whether relationships are linear or non-linear and whether there are interactions between   covariates.

(ref:ch4-rels-bitt) **Multipanel scatterplot of male response distance (cm) and standard length (mm) of European bitterling either without or receiving supplementary feeding with a line of best fit plotted.**  

```{r ch4-rels-bitt, fig.cap='(ref:ch4-rels-bitt)', fig.align='center', fig.dim=c(6, 4), cache = TRUE,  message=FALSE, echo=FALSE, warning=FALSE}

label_supp <- c("No" = "No supplement", 
                "Yes" = "Food supplement")
bitt %>% 
  ggplot(aes(x = sl, y = resp_dist, size = 1)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2, colour = "black") +
  facet_grid(.~ fSupp, 
             scales = "fixed", space = "fixed", 
             labeller=labeller (fSupp = label_supp)) +
  xlab("Male standard length (mm)") + 
  ylab("Male response distance (cm)") +
  My_theme + theme(legend.position = "none") +
  theme(strip.text = element_text(size = 12, face="italic")) +
  theme(panel.border = element_rect(colour = "black", fill=NA, size = 1))
```

The plot of the data in Fig. \@ref(fig:ch4-rels-bitt) does not indicate a non-linear pattern in the data. However, fitted lines for the relationship between male response distance (cm) and standard length (mm) do suggest that the nature of this relationship may vary with feeding treatment, implying a potential interaction between fish size and feeding treatment; i.e. the slopes differ between treatments. An interaction would mean that the relationship between response distance and standard length depends on nutritional state. Interactions like this one are biologically interesting. Given the pattern in these data, inclusion of an interaction term in the model is justified.  

#### Independence of the response variable  

An assumption for a (G)LM is that each observation in a dataset is independent of all others. In the case of the present study each row of data was a different male bitterling. The study was conducted over a short period (10 days) at the peak of the spawning season of the species in a single lake, which reduced the risk of any strong temporal and spatial effects. Observations were also made by a single experimenter, limiting the risk of dependency in the data due to variation in observer bias. On this basis, we will assume the data are independent. 

### Select and fit a statistical model  

The study was designed specifically to understand the extent to which the territorial behaviour of male European bitterling is a function of male size and nutritional state. The dependent variable is male response distance, which the data exploration showed to be continuous and approximately normally distributed (Fig. \@ref(fig:ch4-freqpoly)). There are no zeros in the response variable and there is good reason to believe data are independent. The relationship between male standard length and response distance is approximately linear, irrespective of food supplementation (Fig. \@ref(fig:ch4-rels-bitt)).  

Given these findings from the data exploration, a Gaussian is an appropriate distribution as a starting point. Two covariates will be included in the model; male standard length (continuous) and food supplementation (categorical, with two levels) as well as their interaction, which means the model will have five parameters; an intercept ($\beta_1$), a slope for standard length ($\beta_2$), food supplementation ($\beta_3$), and the interaction between standard length and food supplementation ($\beta_4$), and the variance ($\sigma^2$) of the normal distribution for male response distance.  

The model will include both main terms and an interaction, fitted using the `glmmTMB` package. The model (`Gaus1`) is fitted as: 

```{r ch4-Gaus1, message=FALSE, echo=TRUE, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=50)}
Gaus1 <- glmmTMB(resp_dist ~ sl * fSupp,   
                             data = bitt, 
                             family = gaussian)
```

Before obtaining a summary of the model output, we first conduct model checks.

### Conduct model checks  

Start by using the `simulateResiduals` command from the `DHARMa` package to simulate scaled residuals from the fitted model. These can then be plotted against model fitted values to ensure model `Gaus1` is not misspecified.  

```{r ch4-simRes, message=FALSE, echo=TRUE, warning=FALSE}
Res1 <- simulateResiduals(fittedModel = Gaus1, plot = F)
```

We start by plotting simulated standardised residuals against model fitted values:  

```{r ch4-plotRes, message=FALSE, echo=TRUE, warning=FALSE}
par(mfrow = c(1,1), mar = c(2,2,2,2))
plotResiduals(Res1)
```
  
  
The `plotResiduals` function calculates a quantile regression of the residuals, for the 0.25, 0.5 and 0.75 quantiles. The residuals should be uniformly distributed for a correctly specified model, so the expectation for these regressions are straight lines at 0.25, 0.5 and 0.75, which are displayed as dashed black lines on the plot. In this case, the residuals plot raises no concerns.

We further plot model residuals separately against each covariate in the model:  

```{r ch4-plotResCov, message=FALSE, echo=TRUE, warning=FALSE}
plotResiduals(Res1, form = bitt$sl)
plotResiduals(Res1, form = bitt$fSupp)
```

Again, these plots raise no concerns. Finally, we check the normality of the residuals, dispersion and presence of outliers:  

```{r ch4-plotQQ, message=FALSE, echo=TRUE, warning=FALSE}
plotQQunif(Res1)
```
  
These checks further indicate no problems with the model.  

Should we consider model selection? Is there any justification? Use manual backward selection with `drop1`:  

```{r ch4-drop1, message=FALSE, echo=TRUE, warning=FALSE}
round(drop1(Gaus1),1)
```

The model fit without the interaction (AIC = 408.7) is not substantially different to a model with the interaction (AIC = 409.5), so we could consider dropping this interaction term from the final model. Here we choose to present the full model with the interaction retained.  
  
### Interpret and present model output  

Obtain model output with:  

```{r ch4-sumGaus1, message=FALSE, echo=TRUE, warning=FALSE}
out <- summary(Gaus1)
print(out, digits = 3, signif.stars=FALSE)
```

Table 4.1: **Mean estimates for aggressive response distances (cm) of male European bitterling (_Rhodeus amarus_) as a function of male standard length (mm) and a supplementary feeding treatment, modelled using a Gaussian LM.**  

|Model parameter                             |Estimate|Z value|P value|
|:-------------------------------------------|:------:|:-----:|:-----:|
|Intercept                                   | 47.72  |2.65   |  0.008|
|Standard length                             | 1.98   |5.86   | <0.001|
|Supplementary feeding(yes)                  | 59.14  |2.14   |  0.033|
|Standard length :                           |        |       |       |
|Supplementary feeding(yes)                  |-0.58   |-1.09  |  0.274|  

These results show a statistically significant positive effect of male bitterling standard length on response distance, with larger males initiating attacks on a rival at greater distances than smaller males (P <0.001). The effect of supplementary feeding for 6-days prior to testing was similarly to increase the aggressive response distance to a rival (P = 0.033). However, there was no evidence for an interaction between male size and supplementary feeding (P = 0.274).  

### Visualise the results  

We can plot the model using the `plot_model` command from the `sjPlot` package.  

(ref:ch4-plotGaus1) **Mean aggressive response distance (cm) of male European bitterling (_Rhodeus amarus_) as a function of male standard length (mm) with and without supplementary feeding, modelled using a Gaussian LM. Shaded areas are 95% confidence intervals. Points are observed data for different males.**  

```{r ch4-plotGaus1, fig.cap='(ref:ch4-plotGaus1)', fig.align ='center', fig.dim = c(5,4), message=FALSE, echo=TRUE, warning=FALSE}
  set_theme(legend.pos = "top")
plot_model(Gaus1, type = "pred", 
                 terms = c("sl", "fSupp"),
                colors = c("firebrick2", "blue"),
             show.data = TRUE,
                 title = "",
           show.legend = T,
          legend.title = "Supp feed",
            axis.title = c("SL (mm)",
                           "Response distance (cm)")) + 
              My_theme

```
<br>  
<br>  
The results of this statistical analysis can be summarised as follows:  
  
  
_A  Gaussian LM was fitted to data to model the aggressive response distance (in cm) of a group of 48 territorial male European male bitterling_ (Rhodeus amarus) _to a model rival. There was a statistically significant positive effect of male standard length (in mm) and supplementary feeding on response distance (P <0.001) (Fig. \@ref(fig:ch4-plotGaus1)). The effect of supplementary feeding for six days prior to testing was to significantly increase male response distance (P = 0.033) (Fig. \@ref(fig:ch4-plotGaus1)). There was no evidence for a significant interaction between male standard length and supplementary feeding on male response distance (P = 0.274)._  

